The role of $4f$ electrons in the stopping power of hafnium

### 2020-02-21 hafnio_ale.tex
include theoretical and conclusion corrections

%=============================================================================================================================================================================================
%=============================================================================================================================================================================================
\begin{abstract}
The stopping power of protons through Hf foil has been studied both experimentally and theoretically. 
The measurements were performed at the Laboratory of Accelerators and X-Ray Diffraction in Lisbon, by using the transmission method on self-supporting stopping material, for (0.6-2.5) MeV protons.
The theoretical developments involved fully relativistic atomic structure calculations for Hf, which required the solution of the Dirac equation. 
The shell-wise local plasma approximation (SLPA) was used to describe the energy transferred to the bound 1s-4f electrons, and the outer four electrons were considered as a free electron gas (FEG). 
 We found the relativistic description of the 4f-shell, and the screening between $4f$ and $5p$ electrons are decisive around the stopping maximum. 
The present theoretical and experimental results have very good agreement in the energy region of the new measurements. 
However, our theoretical stopping cross sections show substantial differences with the semi-empirical models, such as SRIM2013 and ICRU-49, at intermediate to low energies. 
Our calculations suggest the stopping maximum to be larger, and shifted to lower energies, respect to these semi-empirical models. Future measurements below 100 keV would be required to validate our predictions.
\end{abstract}

@@

In this study, we report experimental stopping power cross section values over the incident energy range (0.6-2.5) MeV for protons crossing self-supported Hf thin-film by using the transmission method.
We aim not only to upgrade stopping power data compilations \cite{HPaul03,mondim17} but also to provide useful information about the processes governing the slowing down of protons in multi-electronic targets. 
In the rare earth metals, the $4f$ electrons play a relevant role in the stopping power, being the first shell of bound electrons below the conduction band. As already noted \cite{Roth17}, the free electron gas (FEG) shows unexpected behavior in these elements, which casts doubts on its proper description. With Hf, we found the contribution of the 4f-shell decisive even at impact energies around the stopping maximum, as shown later.

The theoretical approach implemented in this work uses the shell-wise local plasma approximation (SLPA) \cite{mon13} to describe the energy transferred to the bounds 1s-4f electrons, and two different models for the FEG: the screened potential with cusp condition model (SPCC) \cite{mon17}, which is non-linear binary formalism; and the Mermin-Lindhard dielectric formalism (ML) \cite{Mermin}, for energies around the stopping maximum and above. 
Our model requires the relativistic description of the wave functions and binding energies of Hf \cite{mendez2019}, where we consider 4 electrons per atom in the FEG. 
The 14 electrons occupying the 4f subshell of Hf makes up the main contribution to the stopping below the FEG, which causes the cross sections to be very sensitive to a good description of this shell. 
The screening between the 4f and 5p electrons has been considered, and it has been found to play a major role within the SLPA predictions.

@@

Hafnium ($Z=72$, [Xe] $4f^{14} \ 6s^2 \ 5d^2$) belongs to the first group of transition metals, with  4 electrons as FEG (theoretical $r_S=2.14$ a.u.) and $1s-4f$ electrons bound. We compared the computed $r_S$ with the experimental value obtained from the measured energy loss function by Lynch \textit{et al} \cite{lynch75}. The experimental plasmon energy of Hf is $\hbar\omega_P \approx 15.8$ eV, with a width at half maximum of $\delta \approx 4.4$ eV, and $r_S \approx 2.07$ a.u. \cite{lynch75}. The difference of less than $5\%$ between theoretical and experimental $r_S$ assess Hf as a canonical target\cite{mon17}.

@@

For the stopping power due to bound electrons, the SLPA \cite{mon17,mon13} is employed. It is worth mentioning that the only inputs for the SLPA are the space-dependent densities of each shell in the ground state, and their binding energies. Collective processes and screening among electrons are included. Hafnium is a relativistic target, therefore the wave functions and binding energies must be obtained as solutions of the Dirac equation\cite{mendez2019}.

To asses the importance of the relativistic description of the bound electrons, Figure \ref{Binding_E} (a) shows non relativistic and relativistic binding energies $E_{nl\pm}$, where $\pm=j\pm1/2$.
We compare our results with experimental measurements on solid Hf \cite{williams1995} from $1s$ to $4f_{\pm}$. We notice that not only the most inner shells require relativistic calculations, but also the outer $5p$ and $4f$ shells. Figure \ref{Binding_E} (b) presents the relative error of non relativistic and relativistic energies with respect to the experimental data in logarithmic scale. This figure shows very clearly the disability of non-relativistic calculations to describe the experimental data, which surprisingly worsens from the inner to the outer shells. 
%It is clear in this figure the disability of non-relativistic binding energies to describe the experimental data, which surprisingly worsens from the inner to the outer-shells.
%We used here the theoretical binding energies and wave functions of Hf calculated in \cite{mendez2019}. 
%The present theoretical description of Hf considers a FEG of $r_S=2.14$ and 1s-4f electrons as bound.

%The relativistic results for binding energies and wave functions in \cite{mendez2019} 
Our relativistic binding energies present spin-orbit split. However, in total stopping power where the initial state of the excited electron is not measured, the quantum uncertainty in energy $\Delta E$ melts this split. The criteria $\Delta E\Delta t \geq \hbar /2$ merges the energies  $E_{nl+} - E_{nl-}$ for sufficiently small values of $\Delta t$ (the collisional mean-time). In fact, at sufficiently high impact velocity we can expect all target electrons to respond together to the ion passage \cite{lindhard53,chu72}. As in \cite{mon09}, the collisional time was estimated as follows $\Delta t \approx \langle r_i\rangle/v$, with $\langle r_i\rangle$ and $v$ being the orbital mean radio and impact velocity, respectively.

%--------------------------------------------------------------------------------
\begin{figure*}[!t]
\centering
\includegraphics[width=13.cm]{Fig02.eps}
\caption{Theoretical total stopping cross sections of protons in afnium adding FEG and bound $1s$-$4f$ contributions, as mentioned in the inset. The vertical grey dashed-line indicates the energy above which plasmon excitation is possible. %The FEG calculations employ the SPCC model at low energies (no plasmon excitation) and the ML formalism (includes plasmon excitation). 
The SLPA calculations for $1s$-$4f$ bound electrons with and without $5p-4f$ screening give different total stopping values, displayed with solid and dotted lines, respectively.}
\label{slpa4f}
\end{figure*}
%========================================================================================

The SLPA calculates the contribution of each subshell of bound electrons to the total stopping cross sections. We found that for every sub-shell of Hf, at the impact energy at which this sub-shell began to contribute, the spin-orbit split was unresolved.  Then the $nl$-electrons should be considered together, responding to the ion passage as a single gas of electrons with density $\delta_{nl}(r)$ and a mean binding energy $E_{nl}$. This is important within the SLPA calculations because of the screening among electrons of the \textit{same} binding energy. For example, the $4f_{-}$ and $4f_{+}$ of Hf can only be resolved for impact energies $E<0.05$ keV, but the contribution of $4f$ to the total stopping is negligible if $E<40$ keV. Moreover, the $5p$ and $4f$ electrons of Hf are very close in energy ( $\Delta E_{5p-4f} \approx 1 a.u.$ \cite{mendez2019}) %merges them 
and react together for impact energies $E>40$ keV (inter-shell screening).

@@

In figure \ref{slpa4f} we display the present theoretical stopping cross section of Hf for protons with and without the $5p-4f$ screening. We added the FEG and the bound electron contributions as explained above. The minimum energy for plasmon excitation was estimated as $37$ keV. We used the non-perturbative SPCC model for impact energies $E \leq 37$ keV, and the perturbative ML calculation above this energy. %The range of validity of the perturbative model below 100 keV is arguable, however the asymetric $Z_{ion}/Z_{target}$ relation is clear. 
%The importance of the 4f-5p screening is signed with arrows in this figure. 
Clearly, considering $5p-4f$ electrons as a single group of 20 electrons with screening among them gives lower stopping values than the addition of the separate $5p$ and $4f$ contributions. This is a shell correction that can only be considered within a many electron model such as the SLPA.

@@

The present data and theoretical results are displayed in table \ref{table01}, and figure \ref{F03}. As can be seen in Table~\ref{table01}, an overall relative uncertainty of around 5$\%$ was achieved for the experimental stopping power values, mainly due to the uncertainty in the hafnium foil thickness. %A good agreement in the whole energy range is found between our new data and the previous measurements made by Sirotinin \textit{et al.} \cite{Sirotinin}, as depicted in Fig.~\ref{F03}. Also shown in the same figure are our theoretical stopping cross sections together with the semi-empirical SRIM-2013 curve \cite{Ziegler01} and the suggested values by the ICRU-49 report \cite{ICRU49}.

@@

There is good agreement between the present theoretical results, the new data in Table~\ref{table01}, and also with previous data by Sirotinin \cite{Sirotinin}, except for the lowest energy measurement at 80 keV. It is interesting that our full theoretical curve differs from SRIM-2013 one for impact energies below 100 keV. We obtain a stopping maximum around 40 $10^{-15}$ eV cm$^2$/atom  at 65 keV. Instead, following the up-to-now only set of data \cite{Sirotinin}, SRIM-2013 suggests a lower stopping maximum %of 37.4 $10^{-15}$ eV cm$^2$/atom 
at impact energy of 115 kev. It is worth to mention that similar theoretical results using the experimental $r_S=2.07$ a.u. instead of the theoretical $r_S=2.14$ a.u. give the stopping maximum at the same impact energy but $4\%$ higher. Future experiments for
impact energies below 100 keV would be important to complete this study.

%=============================================================================================================================================================================================
%=============================================================================================================================================================================================
\section{Conclusion}
\label{conclusion}
In this work, we have used the transmission method to experimentally determine stopping power cross section values for (0.6-2.5) MeV protons incident on self-supporting Hf foils with an overall uncertainty of around 5\%. Additionally, we calculated values extracted from the theoretical framework that involved the relativistic wave functions and binding energies of Hf, and considered 4 electrons per atom in the free electron gas. The shell-wise local plasma approximation was employed to describe the energy transferred to the bound 1s-4f electrons, and two different models for the FEG: the screened potential with cusp condition (SPCC model) for energies below that of the plasmon excitation, and the Mermin-Lindhard dielectric formalism, for energies around the stopping maximum and above. Present theoretical stopping cross sections cover an extensive energy range from 1 keV/amu-10 meV/amu.

%Our results have been confronted to both experimental values found in the literature and semi-empirical results. 
At high impact energies, the new stopping  measurements are in good agreement with our theoretical results, and also with already published experimental data and semi-empirical values calculated by means of SRIM-2013 and ICRU-49.  However, we call the attention that around the stopping maximum and at lower impact energies the difference between our full-theoretical results and SRIM is substantial. To our knowledge, these are the first theoretical calculations of stopping in Hf taking into account the atomic relativistic effect.
%for energies around and below the stopping maximum, the present full-theoretical results clearly differ from semiempirical SRIM values, which are based on previous data. 
Future experiments for impact energies below 100 keV would be important to complete this study.
%=======================================================================================================================================
\begin{acknowledgments}
This work was funded by VRAC Grant Number L1-17 of Universidad Tecnol\'ogica Metropolitana, Chile; also by the Consejo Nacional de Investigaciones Científicas y Técnicas (CONICET), the Agencia Nacional de Promoción Científica y Tecnológica (ANPCyT), and Universidad de Buenos Aires (UBA), of Argentina.

>>>>

%=============================================================================================================================================================================================
%=============================================================================================================================================================================================
\begin{abstract}
The stopping power of protons through Hf foil has been studied both experimentally and theoretically. 
The measurements were performed at the Laboratory of Accelerators and X-Ray Diffraction in Lisbon, by using the transmission method on self-supporting stopping material, for (0.6-2.5) MeV protons.
The theoretical developments involved fully relativistic atomic structure calculations for Hf, which required the solution of the Dirac equation. 
The shell-wise local plasma approximation (SLPA) was used to describe the energy transferred to the bound $1s$-$4f$ electrons, and the outer four electrons were considered as a free electron gas (FEG). 
 We found the relativistic description of the $4f$-shell, and the screening between $4f$ and $5p$ electrons are decisive around the stopping maximum. 
The present theoretical and experimental results have very good agreement in the energy region of the new measurements. 
However, our theoretical stopping cross sections show substantial differences with the semi-empirical models, such as SRIM2013 and ICRU-49, at intermediate to low energies. 
Our calculations suggest the stopping maximum to be larger, and shifted to lower energies, respect to these semi-empirical models. Future measurements below 100 keV would be required to validate our predictions.
\end{abstract}

@@

In this study, we report experimental stopping power cross section values over the incident energy range (0.6-2.5) MeV for protons crossing self-supported Hf thin-film by using the transmission method.
We aim not only to upgrade stopping power data compilations \cite{HPaul03,mondim17} but also to provide useful information about the processes governing the slowing down of protons in multi-electronic targets. 
In the rare earth metals, the $4f$ electrons play a relevant role in the stopping power, being the first shell of bound electrons below the conduction band. As already noted \cite{Roth17}, the free electron gas (FEG) shows unexpected behavior in these elements, which casts doubts on its proper description. With Hf, we found the contribution of the $4f$-shell decisive even at impact energies around the stopping maximum, as shown later.

The theoretical approach implemented in this work uses the shell-wise local plasma approximation (SLPA) \cite{mon13} to describe the energy transferred to the bounds $1s$-$4f$ electrons, and two different models for the FEG: the screened potential with cusp condition model (SPCC) \cite{mon17}, which is non-linear binary formalism; and the Mermin-Lindhard dielectric formalism (ML) \cite{Mermin}, for energies around the stopping maximum and above. 
Our model requires the relativistic description of the wave functions and binding energies of Hf \cite{mendez2019}, where we consider 4 electrons per atom in the FEG. 
The 14 electrons occupying the $4f$ subshell of Hf makes up the main contribution to the stopping below the FEG, which causes the cross sections to be very sensitive to a good description of this shell. 
The screening between the $4f$ and $5p$ electrons has been considered, and it has been found to play a major role within the SLPA predictions.

@@

Hafnium ($Z=72$, [Xe] $4f^{14} \ 6s^2 \ 5d^2$) belongs to the first group of transition metals, with  4 electrons as FEG (theoretical $r_S=2.14$ a.u.) and $1s$-$4f$ electrons bound. We compared the computed $r_S$ with the experimental value obtained from the measured energy loss function by Lynch \textit{et al} \cite{lynch75}. The experimental plasmon energy of Hf is $\hbar\omega_P \approx 15.8$ eV, with a width at half maximum of $\delta \approx 4.4$ eV, and $r_S \approx 2.07$ a.u. \cite{lynch75}. The difference of less than $5\%$ between theoretical and experimental $r_S$ assess Hf as a canonical target\cite{mon17}.

@@

For the stopping power due to bound electrons, the SLPA \cite{mon17,mon13} is employed. It is worth mentioning that the only inputs for the SLPA are the space-dependent densities of each shell in the ground state, and their binding energies. Collective processes and screening among electrons are included. Hafnium is a relativistic target, therefore the wave functions and binding energies must be obtained as solutions of the Dirac equation.

To assess the importance of the description of the bound electrons, Fig. \ref{Binding_E} shows the (a) non relativistic and relativistic binding energies\cite{mendez2019}  $E_{nl\pm}$, where $\pm=j\pm1/2$, and (b) their relative errors with respect to experimental measurements on solid Hf \cite{williams1995}, from $1s$ to $4f_{\pm}$. 
We notice that not only the most inner shells require relativistic calculations, but also the outer $5p$ and $4f$ shells. 
%Figure \ref{Binding_E} (b) presents the relative error of non relativistic and relativistic energies with respect to the experimental data. 
Furthermore, this figure shows very clearly the disability of non-relativistic calculations to describe the experimental data, which surprisingly worsens from the inner to the outer shells.

Our relativistic binding energies present spin-orbit split. 
However, the quantum uncertainty in energy $\Delta E$ melts this split in total stopping power, where the initial state of the excited electron is not measured. 
The criteria $\Delta E\Delta t \geq \hbar /2$ merges the energies  $E_{nl+} - E_{nl-}$ for sufficiently small values of $\Delta t$ (the collisional mean-time). 
In fact, at sufficiently high impact velocity, we can expect all target electrons to respond together to the ion passage \cite{lindhard53,chu72}. 
Following previous work\cite{mon09}, the collisional time is estimated as $\Delta t \approx \langle r_i\rangle/v$, with $\langle r_i\rangle$ and $v$ being the orbital mean radio and impact velocity, respectively.

%--------------------------------------------------------------------------------
\begin{figure*}[!t]
\centering
\includegraphics[width=13.cm]{Fig02.eps}
\caption{Theoretical total stopping cross sections of protons in hafnium adding FEG and bound $1s$-$4f$ contributions, as mentioned in the inset. The vertical grey dashed-line indicates the energy above which plasmon excitation is possible. %The FEG calculations employ the SPCC model at low energies (no plasmon excitation) and the ML formalism (includes plasmon excitation). 
The SLPA total stopping values for $1s$-$4f$ bound electrons with and without $5p$-$4f$ screening are given with solid and dotted lines, respectively.}
\label{slpa4f}
\end{figure*}
%========================================================================================

The SLPA calculates the contribution of each subshell of bound electrons to the total stopping cross sections. We found that for every sub-shell of Hf, at the impact energy at which this sub-shell began to contribute, the spin-orbit split was unresolved.  
Then, the $nl$-electrons should be considered together, responding to the ion passage as a single gas of electrons with density $\delta_{nl}(r)$ and a mean binding energy $E_{nl}$. 
This feature is important within the SLPA calculations because it accounts for the screening among electrons of the \textit{same} binding energy. For example, the $4f_{-}$ and $4f_{+}$ of Hf can only be resolved for impact energies $E<0.05$ keV, but the contribution of $4f$ to the total stopping is negligible for $E<40$ keV. Moreover, the $5p$ and $4f$ electrons of Hf are very close in energy ( $\Delta E_{5p-4f} \approx 1 $ a.u.\cite{mendez2019}) 
and they react together for impact energies $E>40$ keV (inter-shell screening).

@@

In Fig. \ref{slpa4f}, we display the present theoretical stopping cross section of Hf for protons with and without the $5p$-$4f$ screening. We added the FEG and the bound electron contributions, as explained above. The minimum energy for plasmon excitation was estimated to $37$ keV. We used the non-perturbative SPCC model for impact energies $E \leq 37$ keV, and the perturbative ML calculation above this energy. %The range of validity of the perturbative model below 100 keV is arguable, however the asymetric $Z_{ion}/Z_{target}$ relation is clear. 
%The importance of the 4f-5p screening is signed with arrows in this figure. 
Clearly, considering $5p$-$4f$ electrons as a single group of 20 electrons with screening among them gives lower stopping values than the addition of the separate $5p$ and $4f$ contributions. Notice that this shell correction can only be considered within a many electron model such as the SLPA.

@@

The present data and theoretical results are displayed in Table \ref{table01} and Fig. \ref{F03}. As can be seen in Table~\ref{table01}, an overall relative uncertainty of around 5$\%$ was achieved for the experimental stopping power values, which are mainly due to the uncertainty in the hafnium foil thickness. %A good agreement in the whole energy range is found between our new data and the previous measurements made by Sirotinin \textit{et al.} \cite{Sirotinin}, as depicted in Fig.~\ref{F03}. Also shown in the same figure are our theoretical stopping cross sections together with the semi-empirical SRIM-2013 curve \cite{Ziegler01} and the suggested values by the ICRU-49 report \cite{ICRU49}.

@@

In Fig. \ref{F03}, we have good agreement between the present theoretical results and the measurements given in Table~\ref{table01}. Our theoretical approach also agrees with previous data by Sirotinin \cite{Sirotinin}, except for the lowest energy measurement at 80 keV. It is interesting that our full theoretical curve differs from SRIM-2013 for impact energies below 100 keV. We obtain a stopping maximum around 40 $10^{-15}$ eV cm$^2$/atom  at 65 keV. Instead, following the up-to-now only set of data \cite{Sirotinin}, SRIM-2013 suggests a lower stopping maximum %of 37.4 $10^{-15}$ eV cm$^2$/atom 
at impact energy of 115 kev. It is worth mentioning that similar theoretical results, using the experimental $r_S=2.07$ a.u. instead of the theoretical $r_S=2.14$ a.u., give the stopping maximum at the same impact energy but $4\%$ higher. Future experiments for impact energies below 100 keV would be important to complete this study.

%=============================================================================================================================================================================================
%=============================================================================================================================================================================================
\section{Conclusion}
\label{conclusion}
In this work, we have used the transmission method to experimentally determine stopping power cross section values for (0.6-2.5) MeV protons incident on self-supporting Hf foils with an overall uncertainty of around 5\%. Additionally, we calculated values extracted from the theoretical framework that involved the relativistic wave functions and binding energies of Hf, and considered 4 electrons per atom in the free electron gas. The shell-wise local plasma approximation was employed to describe the energy transferred to the bound $1s$-$4f$ electrons, and two different models for the FEG: the screened potential with cusp condition (SPCC model) for energies below that of the plasmon excitation, and the Mermin-Lindhard dielectric formalism for energies around the stopping maximum and above. Present theoretical stopping cross sections cover an extensive energy range from 1 keV/amu-10 meV/amu.

At high impact energies, the new stopping measurements are in good agreement with our theoretical results. We also agree with previously published experimental data and semi-empirical values calculated by means of SRIM-2013 and ICRU-49.  However, we call the attention that around the stopping maximum and at lower impact energies the difference between our full-theoretical results and SRIM is substantial. To the best of our knowledge, these are the first theoretical calculations of stopping in Hf taking into account the atomic relativistic effect.
Future experiments for impact energies below 100 keV would be required to validate the predictions of our theoretical model.
%=======================================================================================================================================
\begin{acknowledgments}
This work was funded by VRAC Grant Number L1-17 of Universidad Tecnol\'ogica Metropolitana, Chile; also by the Consejo Nacional de Investigaciones Cient\'ificas y T\'ecnicas (CONICET), the Agencia Nacional de Promoci\'on Cient\'ifica y Tecnol\'ogica (ANPCyT), and Universidad de Buenos Aires (UBA), from Argentina.
